A Handbook for Tenured Track Faculty in Computer Science in the US

### 2024-07-08 faculty.tex
updates

\subsection{Serving on Program Committees (PCs) and Editorial Boards} You should try to serve on the PCs  of the \emph{top} conferences in your field. This is a good way to know how to write good papers, and get to know the literature, what is hot in your field, and the people (who do what). It is also a good way to get your name out.  Similarly to PCs, you should also try to serve on the editorial boards of top journals in your field.

Serving on PC (or editor boards) essentially means reviewing papers. You will have to read papers and write reviews, and then argue for why a paper should be accepted or not.  You will also have to discuss with other PC members (or editors) and come to a consensus.  Typically the review process can last for several weeks, and there are also rebuttals and conditionally acceptances (which you should already be familiar with as authors).

@@

\subsection{Personal Experience} I mostly cold email people for opportunities. For example, I directly email OOPSLA and PLDI PC chairs and asked, and they invited. I tried this with ICSE (and also nominate myself through the official form), and it never works, even though I have many papers at ICSE (that leaves a bad taste in my mouth about ICSE, but that is another story).  I also got invited to serve on a PC because a senior colleague recommended me.
For NSF, I just emailed program managers and asked to serve on their panels, and they invited me.

@@

\section{Service to the Department}
Your department chair will assign you to some committees to help run the department.  For tenure-track faculty the services are often lighter than tenured ones (e.g., less leadership roles).

@@

\subsection{Personal Experience} Since coming to GMU, I have been very active in serving the department.  In addition to the typical services load, I volunteer to serve in additional committees and leadership roles. For example, I maintain the CSRankings DB for the department, organize the Open House to recruit admitted students, participate in the Dept Chair renewal committee, etc.
Even though I am not on the search committee, I meet with every faculty candidate (I talked to 22 people in Spring'23, which was during my study break). I also agree to be director of the MS SWE program, which is a lot of work and typically done by tenured faculty.

Occasionally I feel overwhelmed by having many services, but I also feel that I am making a difference in the department and that is rewarding.
I am also building a good relationship with my colleagues (and also head-butt with some), and that is important for my tenure case. But of course I don't do this because of my tenure case, I do it because I enjoy it and believe I can make a difference.
In addition, I learn a lot from being in various roles (e.g., how class scheduling work, how the department is run, how to recruit students and recruit faculty, how to deal with difficult people and situation, etc).

In general, I was way more active than most junior faculty in the department.  I was also a lot more involved in the department than I was in my previous department at UNL. Here I share the vision of the department and want to help achieve it.
In short, if you want to make a difference, and \emph{feel comfortable} at handling additional load, you can volunteer for more services. Otherwise, you can just ``stay low'',
do the required services and focus on your research and teaching. This is probably the best strategy for most junior faculty.

@@

\subsection{Personal Experience} I have limited experience with these services. I served in a couple of grants and scholarship committees, in which I review applications for scholarship and grants for students. I also help represent the college in serving as judge of the Kimmy-Duong foundation scholarship. These are time-consuming (e.g., the KimmyDuong scholarship reviews and interview process happen during weekends!), but they are also rewarding as you get to know students and help them (many have unique and interesting stories).

@@

\chapter{Renewal and Tenure Application}

Toward the end of your 3rd year (or early 4th year), you will need to submit a \emph{renewal} application.  The department will then review your application and vote on whether to renew you for your 4th (and perhaps also 5th year). At some university this renewal is almost guaranteed, but at some other universities, it is not and people do get denied.

Then, after 5 or 6 years, it is time for you to go up (or apply) for tenure.  You will need to submit a tenure package.  The department will then review your package and vote on whether to recommend you for tenure.  If the department votes in your favor, the package will then go to the college, and then to the university (e.g., provost, president, board of visitors, etc).  The university will then make the final decision on whether to grant you tenure (typically they follow the department's recommendation, but not always).

I will focus on the tenure application process, as the renewal process is similar but less intense (e.g., no external letters).

@@

\begin{itemize}
  \item Fall 2021: Moved to GMU with accelerated tenure clock (meaning I don't have to restart from the beginning as I have already been at UNL since 2016).
  \item (late) Fall 2022: Told my dept. chair that I plan to apply for tenure in 2023.
  \item Spring 2023: Prepared my tenure package, which includes writing statements and think reference letters.  Also took this semester off (no teaching no services) because GMU has a pre-tenure study break policy.
  \item Summer 2023: Submitted my tenure package.
  \item Fall 2024: Gave a tenure talk. Department voted in my favor, and then the college voted in favor, and the decision goes up to the university.
  \item Spring 2024: In May, got the official notification that I got tenure and promoted to Associate Prof., beginning in Fall 2024.
\end{itemize}

@@

Here is my \href{https://dynaroars.github.io/people/nguyenthanhvuh/files/pt/slides.pdf}{talk}. I focus on my recent research direction on neural network verification and just briefly on my past work. Overall, I didn't have to spend too much time on preparing because I reused slides from my job and research talk and just added at the end a ``stats'' slide on my funding, pubs, teaching, services, etc.

@@

You will need a senior faculty member (or ``champion'') from your department to present your case in the department P\&T meeting. This person will present your achievements and argue for you.  Typically, this person is your mentor, but it can be anyone in the department that you trust and who knows your work well.
A good champion won't likely be able to help a bad case, but a bad champion can hurt a good case.  So do pay attention to this.

At GMU I was assigned a mentor, Paul Ammann, right when I joined.  For my tenure case, he put together my ``stats'' page and asked me to review it. I also discussed with him about letter writers, etc.  Overall, I think my case is relatively strong and so the whole thing was very ``smooth sail'' according to Paul. However, I did get much advice and guidance from him in other things e.g., how to run the MS SWE program (more on that later).

\begin{commentbox}[Vu:]
    While I have had not experienced with an ineffective mentor, I have heard about cases where the mentor did not do a good job, which can hurt strong candidates during the 4-th year renewal process. For example, a candidate might have few, but impactful high-quality publications, and the mentor did not emphasize this enough. In some case, an ineffective mentor actually turn a significant accomplishment into a negative, e.g., getting a 10-year Most Influential paper award is a big deal in certain fields, but the mentor did not emphasize this or allow it being downplayed by others (and allow idiotic comments from other faculty such as the work has been done 10 year ago). So do pay attention to this and make sure you have a good mentor who can defend you. If necessary, you can ask your chair to assign you a different mentor.

\end{commentbox}
\subsection{Others}

\paragraph{Tenure Visits} Many people will visit other universities to give talks and meet people.  This is a good way to get your name out and to get letters (e.g., from the person that host you).  Physical visits were not possible during the pandemic, but virtual visits are possible and in fact easier. However,  I did not do any tenure visits or give talks in the year leading up to my tenure application. This was because I am not a travel person and also I felt that it would not help much in my case.

\paragraph{Inviting speakers to colloquium} Many departments have colloquium series where they invite speakers to give talks.  You can invite potential letter writers to your department's colloquium.

In general, these are a good way to get to know people in your field and to get your name out. Many of these, especially tenure visits, are common practices for junior faculty applying for tenure.

\section{Renewal Package}

>>>>

\subsection{Serving on Program Committees (PCs) and Editorial Boards} You should try to serve on the PCs  of the \emph{top} conferences in your field. This is a good way to know how to write good papers, and get to know the literature, what is hot in your field, and the people (who do what). It is also a good way to get your name out.

You should also try to serve on the editorial boards of top journals in your field.
Similar to being in conf. PCs, this essentially means reviewing paper submissions. However, unlike PC reviews, you typically will not have to discuss and reach a consensus with other reviewers. The journal editor will make the decision based on the reviews.

@@

\paragraph{Personal Experience} I mostly cold email people for opportunities. For example, I directly email OOPSLA and PLDI PC chairs and asked, and they invited. I tried this with ICSE (and also nominate myself through the official form), and it never works, even though I have many papers at ICSE (that leaves a bad taste in my mouth about ICSE, but that is another story).  I also got invited to serve on a PC because a senior colleague recommended me.
For NSF, I just emailed program managers and asked to serve on their panels, and they invited me.

@@

\section{Service to the Department}\label{sec:services-dept}
Your department chair will assign you to some committees to help run the department.  For tenure-track faculty the services are often lighter than tenured ones (e.g., less leadership roles).

@@

\paragraph{Personal Experience} Since coming to GMU, I have been very active in serving the department.  In addition to the typical services load, I volunteer to serve in additional committees and leadership roles. For example, I maintain the CSRankings DB for the department, organize the Open House to recruit admitted students, participate in the Dept Chair renewal committee, etc.
Even though I am not on the search committee, I meet with every faculty candidate (I talked to 22 people in Spring'23, which was during my study break). I also serve in the Chair Renewal committee and as the director of the MS SWE program, which is a lot of work and typically done by tenured faculty.

Occasionally I do feel overwhelmed by having many services, but I share the vision of the department and want to help achieve it.
I learn a lot from being in various roles (e.g., how class scheduling work, how the department is run, how to recruit students and recruit faculty, how to deal with difficult people and situation, etc).
I am also building a good relationship with my colleagues (and also head-butt with some), and that is important for my tenure case.

In general, I was way more active than most junior faculty in the department.  I was also a lot more involved in the department than I was in my previous department at UNL.
In short, if you want to make a difference, and \emph{feel comfortable} at handling additional load, you can volunteer for more services. Otherwise, you can just ``stay low'',
do the required services and focus on your research and teaching. This is probably the best strategy for most junior faculty.

@@

\paragraph{Personal Experience} I have limited experience with these services. I served in a couple of grants and scholarship committees, in which I review applications for scholarship and grants for students. I also help represent the college in serving as judge of the Kimmy-Duong foundation scholarship. These are time-consuming (e.g., the KimmyDuong scholarship reviews and interview process happen during weekends!), but they are also rewarding as you get to know students and help them (many have unique and interesting stories).

@@

\chapter{Tenure Application}

After 5 or 6 years, it is time for you to go up (or apply) for tenure.  You will need to submit a tenure package.  The department will then review your package and vote on whether to recommend you for tenure.  If the department votes in your favor, the package will then go to the college, and then to the university (e.g., provost, president, board of visitors, etc).  The university will then make the final decision on whether to grant you tenure (typically they follow the department's recommendation, but not always).

@@

\begin{itemize}
  \item Fall 2021: Moved to GMU with accelerated tenure clock (meaning I don't have to restart from the beginning as I have already been at UNL since 2016).
  \item (late) Fall 2022: Told my dept. chair that I plan to apply for tenure in Summer 2023.
  \item Spring 2023: Prepared my tenure package, which includes writing statements and thinking about reference letter writers.  Also took this semester off (no teaching no services) because GMU has a pre-tenure study break policy.
  \item Summer 2023: Submitted my tenure package.
  \item Fall 2024: Gave a tenure talk. Department voted in my favor, and then the college voted in favor, and the decision goes up to the university.
  \item Spring 2024: In May, got the official notification that I got tenure and promoted to Associate Prof., beginning in Fall 2024.
\end{itemize}

@@

Here is my \href{https://dynaroars.github.io/people/nguyenthanhvuh/files/pt/slides.pdf}{talk}, which took place in Fall'23 (after I submit my tenure package in Summer'23). I focus on my recent research direction on neural network verification and just briefly on my past work. Overall, I didn't have to spend too much time on preparing because I reused slides from my job and research talk and just added at the end a ``stats'' slide on my funding, pubs, teaching, services, etc.

@@

You will need a senior faculty member from your department to present your case in the department P\&T meeting. This person (i.e., your ``champion'') will present your achievements and argue for your case.  Typically, the champion is your mentor, but it can be anyone in the department that you trust and who knows your work well.
A good champion won't likely be able to help a bad case, but a bad champion can hurt a good case.  So do pay attention to this.

At GMU I was assigned a mentor when I joined.  I discussed with him about letter writers, and he put together my ``stats'' page and asked me to review it. Overall, I think my case is relatively strong and so the whole thing was a ``smooth sail'' according my mentor.

\begin{commentbox}[Vu:]
    If a case is not strong enough (e.g., lack of funding) then the department chair and mentor will likely advise the candidate against applying for tenure and instead seek for extension to improve. I think this is a good practice to  ``save face'' for the candidate and avoid a negative tenure decision, in which case they likely have to find a new job.
\end{commentbox}

\subsection{Tenure Visits and Colloqium} Close to the tenure application (e.g., during the Spring), some tenure candidates will visit other universities to give talks and meet people. 
Sometimes the candidate is assigned to be the chair of the colloquium in the department, and they can invite their potential tenure letter writers to give talks.
These are all good ways to get your name out and to get letters.
The traditional way is a physical visit, but since the pandemic, many of these visits are done remotely (which are more convenient for many people).

I did not do any tenure visits or host anyone for the purpose of getting tenure letters. This was partly because I am not a travel person, and also I felt that it would not help much in my case; I believe that I have strong record and very good relationship with my colleagues and the dept.  Instead, during the Spring'23 semester before my tenure application, I put lots of effort in helping the university in recruiting faculty (by meeting with 20+ faculty candidates), even though I am not in the search committee and really didn't have service requirements as I am on study break that semester.  As mentioned in \autoref{sec:services-dept}, I didn't really do any of these for my tenure case, but because I enjoy the working env at GMU and want to help out.

\section{3rd year Renewal Package}

%As mentioned, a good dept chair or mentor will likely advise against applying for tenure unless they think your case is strong enough.  This is to avoid a direct tenure deny, which can be very hard and bad for the candidate. 
%On the other hand, 4-th year renewal is unavoidable and often used as a ``warning'' to the candidate that they need to improve.

Toward the end of your 3rd year (or early 4th year), you will need to submit a ``midterm'' \emph{renewal} application to renew your contract for the 4th year (and perhaps also 5th year).
While somewhat similar to tenure evaluation, this is less intense and  does not require reference letters or presentation.
%The department will then review your application and vote on whether to renew you for your 4th (and perhaps also 5th year). 
Your mentor presents and argues for your case in the dept. meeting, but the decision is made within the dept. and usually does not go up to the college or university.  It is still signed by the Dean, who often just follows the dept's recommendation.

While the renewal is straightforward for most people, it can be surprise and shock to some.

%I didn't have a good experience with the 4-th year renewal process at UNL, which is one of the reasons why I had to leave UNL.

% \begin{commentbox}[Vu:]
%     While I have not experienced with an ineffective mentor, I have heard about cases where the mentor did not do a good job, which can hurt strong candidates during the 4-th year renewal process. For example, a candidate might have few, but impactful high-quality publications, and the mentor did not emphasize this enough. In some case, an ineffective mentor actually turn a significant accomplishment into a negative, e.g., getting a 10-year Most Influential paper award is a big deal in certain fields, but the mentor did not emphasize this or allow it being downplayed by others (and allow idiotic comments from other faculty such as the work has been done 10 year ago). So do pay attention to this and make sure you have a good mentor who can defend you. If necessary, you can ask your chair to assign you a different mentor.

% \end{commentbox}
